Order reduction and MILP models for certifying Robustness of DNNs globally.

### 2025-07-31 AAAI2026/introduction.tex
global pass

Deep neural networks (DNNs for short) have demonstrated remarkable capabilities, achieving human-like or even superior performance across a wide range of tasks. However, their robustness is often compromised by their susceptibility to input perturbations \cite{szegedy}. This vulnerability has catalyzed the verification community to develop various methodologies, each presenting a unique balance between completeness and computational efficiency \cite{Marabou,Reluplex,deeppoly}. This surge in innovation has also led to the inception of competitions such as VNNComp \cite{VNNcomp}, which aim to systematically evaluate the performance of neural network verification tools. Among them, NNenum \cite{nnenum}, Marabou \cite{Marabou,Marabou2}, and PyRAT \cite{pyrat} MnBAB \cite{ferrari2022complete}, built upon ERAN \cite{deeppoly} and PRIMA \cite{prima}; and importantly, $\alpha,\beta$-CROWN \cite{crown,xu2020fast}, the winner of the last 4 VNNcomp, benefiting from branch-and-bound based methodology \cite{cutting,BaB}.

The benchmarks usually focus on {\em local} robustness, i.e. given a DNN, an image and a small neighborhood around this image, is it the case that all the images in the neighborhood are classified in the same way by the DNN? The neighborhood is provided by a maximal perturbation of the input image, often an $L_0$ perturbation, i.e. every subpixel of the input image can vary in a very small range, typically 
$\frac{2}{255}$ (that is 2 levels of grey/blue/right/green), as $L_0$ is perfectly linear and perturbations of subpixels are independent. While the $L_0$ norm is the easiest to handle, it is not necessarily the most meaningful perturbation.
More importantly, verification tools for local robustness are hard to be used in a real-time decision making pipeline: considering an autonomous car with a video feed from the dashboard, it is computationally too intensive to certify in few ms images of the feed to e.g. skip non-robust images and only consider certified robust images for the decision-making process.

@@

In this paper, we consider {\em global} robustness, that is we do not restrict the certification process to the neighborhood of an input. We propose a two steps procedure. The first step, performed offline once, 
computes global bounds on the shift between 
output values of different decision classes due to the perturbation. That is, considering decision classes $C$ and $D$ and perturbation $\varepsilon$, compute 
{\em global bound} $\beta^\varepsilon_{C,D} \geq \max_{I,I', |I-I'| \leq \epsilon}(value_{I}(C) - value_{I}(D) + value_{I'}(D) - value_{I'}(C))$, where $value_{J}(X)$ is the output value of class $X \in \{C,D\}$ for input image $J \in \{I,I'\}$. %Notice that as $I,I',C,D$ have symetrical roles, we can choose $\beta^\varepsilon_{D,C} = \beta^\varepsilon_{C,D}$. Also, as $I,I'$ have symetrcal role, the minimum value is exactly $-$ the maximum value. 
%Hence, if we have $n$ decision classes, we only have to compute $n (n-1)/2$ bounds. 
Bound $\beta^\varepsilon_{C,D}$ needs to be computed once for a DNN, 
and it is valid over the whole input space; compared with $k$ computations for {\em local robustness}, once for each of the $k$ input images, with results only valid for these $k$ images.

@@

	\item Since the problem is a very hard verification problem (the values each neuron can take is not bound to a small neighborhood), we follow the accurate MILP technique in \cite{ATVA25}. The number of variables a priori doubles compared with local robustness, from each neuron value in the perturbed image to each neuron value in the perturbed image {\em and} in the original image, as the original image is no more fixed. Recall that the worst case complexity of MILP is exponential in the number of binary variables \cite{DivideAndSlide}. A straightforward MILP model would be to use the {\em classical MILP model} \cite{MILP} for each of these variables. While the full MILP model is exact, it is too computationally intensive to be run for anything but small DNNs. As explained in \cite{ATVA25}, many variables need to be linearly relaxed to reach useable bounds. The main issue with the classical model is that its linear relaxations is extremely inaccurate, as the variables are independent of each other. Instead, we develop another exact MILP model, called the {\em "2v" model}, where the variables are the values of the perturbed neurons, as well as the difference between the original and the perturbed neuron values (called the {\em diff variables}). We study and encode exactly how the diff variables evolve after passing through a ReLU (figure \ref{fig.2v}). In this model, the linear relaxation is much more accurate, as each 
	{\em diff variable} can be bounded 	after a ReLU as a function of the value of the {\em diff variables} before the ReLU. We propose 2 lighter models, namely the {\em "3v" model} which decouples one of the variable (less constraint means less accuracy), as well as an efficient {\em "1v" model} only considering the {\em diff variables} but forgetting the value of the perturbed neurons (with half the (binary) variables).

	\item We adapt the Solution Aware Scoring from \cite{ATVA25} to our novel MILP models, in order to select the most important ReLUs to be treated using complex binary variables, while less important variables are treated using linear relaxation. The chosen number of binary variables depends upon the complexity of the DNN as well as the targeted runtime.

   \item  In terms of perturbations, we consider conjunctions of $L_0$ and $L_1$ norms, which allows to accurately describe perturbations. For instance, "each subpixel is perturbed by at most $\frac{20}{255}$ ($L_0$) and the sum of the absolute value of perturbations over all subpixels is at most $\frac{150}{255}$" ($L_1$ perturbation). While $L_1$ perturbations are not linear (because of the absolute values), reason for which it is seldom used, we explain how to use it as a perturbation in the MILP model without incurring any expansive binary variables (only cheap linear variables are necessary).

	\item Results obtained on the full input space have several shortcomings: while the bounds are fully correct, they are also particularly pessimistic, as all inputs, including absolutely improbable ones very far away from anything the network has been trained on, need to be accounted for. Also, the runtime to obtain the bound is particularly long. Finally, when computing worst-case 
	pairs (image, perturbation), improbable images are generated, hence these  are not meaningful. To solve this problem, we restrict the input space to ignore improbable inputs, using model order reduction techniques from engineering science \cite{aiware}. Specifically, we use Principal Component Analysis (PCA) to reduce the space to a linear input space. We choose the number of dimensions of the space to equal the accuracy of the DNN on the reduced space. 
	%(using a projection to the reduced space then the inverse projection to obtain a very similar image understandable by the DNN). 
	On the MNIST benchmark, this means 20 linear dimensions to match the $97\%$ accuracy of the DNN we considered, instead of the 768 dimensions of the full image space. Using PCA, which is linear, makes it easy to specify perturbations on the actual image (where it is meaningful) rather than on the reduced space.

\item Experimentally, this allows to obtain bounds for a considered pre-trained DNN for MNIST (with accuracy $97\%$) that can certify in real time that $86\%$ of the images are robust for an $L_1$ perturbation of $0.5$, resp. $53\%$ for an $L_1$ perturbation of $1$. We finally experimented our MILP models for {\em regression} tasks, considering a DNN surrogate built from a slow physical finite element model, providing the deformation and the plastic strain of a pipe \cite{aiware}. The DNN has been learnt on a reduced order basis already of dimension 10, from thousands of input dimensions. For regression, the aim is to understand if the DNN surrogate that has been learnt is brittle to small perturbations, which could make it unfaithful to the physical model. Here we want to obtain bounds (which is what our models provide), as well as generating worst-case (image,perturbation) pair, where the strain is particularly different, close to the bound, and that can be run in the physical model to understand whether the DNN surrogate is faithful. Our tool was helpful to provide limited bounds, as well as almost matching worst-case (image,perturbation) pair that was confirmed in the physical model.

\item In both cases, the best bounds were obtained using the "3v" model, as the theoretically more accurate "2v" model did not converge enough even after a very large timeout. Still, when trying to produce worst case pairs (image, perturbation), which is somehow easier than certifying bounds as heuristics to guide towards a solution can be used, the "2v" model found worser-case solutions. Last, the "1v" model produced bounds as good as the "2v" models and much faster than the "3v" models, although slightly larger. That is, all 3 MILP models we propose have favorable trade-offs in different contexts.
\end{enumerate}

>>>>

Deep neural networks (DNNs for short) have demonstrated remarkable capabilities, achieving human-like or even superior performance across a wide range of tasks. However, their robustness is often compromised by their susceptibility to input perturbations \cite{szegedy}. This vulnerability has catalyzed the verification community to develop various methodologies, each presenting a unique balance between completeness and computational efficiency \cite{Marabou,Reluplex,deeppoly}. This surge in innovation has also led to the inception of competitions such as VNNComp \cite{VNNcomp}, which aim to systematically evaluate the performance of neural network verification tools. Among them, NNenum \cite{nnenum}, Marabou \cite{Marabou,Marabou2}, and PyRAT \cite{pyrat} MnBAB \cite{ferrari2022complete}, built upon ERAN \cite{deeppoly} and PRIMA \cite{prima}; and $\alpha,\beta$-CROWN \cite{crown,xu2020fast}, the winner of the last 4 VNNcomp, benefiting from branch-and-bound based methodology \cite{cutting,BaB}.

The benchmarks usually focus on {\em local} robustness, i.e. given a DNN, an image and a small neighborhood around this image, is it the case that all the images in the neighborhood are classified in the same way by the DNN? The neighborhood is provided by a maximal perturbation of the input image, often an $L_\infty$ perturbation, i.e. every subpixel of the input image can vary in a very small range, typically $\frac{2}{255}$ (that is 2 levels of grey/blue/right/green), as $L_\infty$ is perfectly linear and perturbations of subpixels are independent. While the $L_\infty$ norm is the easiest to handle, it is not necessarily the most meaningful perturbation. More importantly, verification tools for local robustness are hard to be used in a real-time decision making pipeline: considering an autonomous car with a video feed from the dashboard, it is computationally too intensive to certify in few ms images of the feed to e.g. skip non-robust images and only consider certified robust images for the decision-making process.

@@

In this paper, we consider {\em global} robustness, that is we do not restrict the certification process to the neighborhood of an input. We follow a two steps procedure. 
The first step, performed offline once, computes global bounds on the shift between 
output values of different decision classes due to the perturbation, as proposed in \cite{vhagar}. 
That is, considering decision classes $C$ and $D$ and perturbation $\varepsilon$, compute 
{\em global bound} $\beta^\varepsilon_{C,D} \geq \max_{I,I', |I-I'| \leq \epsilon}(value_{I}(C) - value_{I}(D) + value_{I'}(D) - value_{I'}(C))$, where $value_{J}(X)$ is the output value of class $X \in \{C,D\}$ for input image $J \in \{I,I'\}$. %Notice that as $I,I',C,D$ have symetrical roles, we can choose $\beta^\varepsilon_{D,C} = \beta^\varepsilon_{C,D}$. Also, as $I,I'$ have symetrcal role, the minimum value is exactly $-$ the maximum value. 
%Hence, if we have $n$ decision classes, we only have to compute $n (n-1)/2$ bounds. 
Bound $\beta^\varepsilon_{C,D}$ needs to be computed once for a DNN, 
and it is valid over the whole input space; compared with $k$ computations for {\em local robustness}, once for each of the $k$ input images, with results only valid for these $k$ images.

@@

	\item We develop a novel exact MILP encoding for the global robustness problem,
	called the {\em "2v" model}, where the variables are the values of the perturbed neurons, as well as the difference between the original and the perturbed neuron values (called the {\em diff variables}). We study and encode exactly in MILP how the {\em diff variables} evolve after passing through a ReLU (Prop.~\ref{Prop2}). Compared with the
	{\em classical MILP model} \cite{MILP} employed in \cite{vhagar,lipshitz},
	the linear relaxation is much more accurate, as each {\em diff variable} can be bounded after a ReLU as a function of the value of the {\em diff variables} before the ReLU, whereas the linear relaxations of the classical model is extremely inaccurate, as the variables are independent of each other. This was witnessed in \cite{lipshitz}, and constraints encoding a part of the linear relaxation of our "2v" model were added explicitly. Experimentally, 
	our "2v" model provides more accurate bounds than competing models, in the same runtime
	(Table \ref{table.classical}).
	%The number of variables a priori doubles compared with local robustness, from each neuron value in the perturbed image to each neuron value in the perturbed image {\em and} in the original image, as the original image is no more fixed. 
	%Recall that the worst case complexity of MILP is exponential in the number of binary variables \cite{DivideAndSlide}. 
	%A straightforward MILP model would be to use the  for each of these variables, as in \cite{vhagar,lipshitz}. The main issue with the classical model is that its linear relaxations is extremely inaccurate, as the variables are independent of each other. Instead, we develop another exact MILP model, 
	Further, we develop two abstract MILP models, namely the {\em "3v" model} which decouples one of the variable, helping the solver, as well as an efficient {\em "1v" model} only considering the {\em diff variables} but forgetting the value of the perturbed neurons.

	%\item We adapt the Solution Aware Scoring from \cite{ATVA25} to our novel MILP models, in order to select the most important ReLUs to be treated using complex binary variables, while less important variables are treated using linear relaxation. The chosen number of binary variables depends upon the complexity of the DNN as well as the targeted runtime.

   \item  In terms of perturbations, we consider conjunctions of $L_\infty$ and $L_1$ norms, which allows to accurately describe perturbations. For instance, "each subpixel is perturbed by at most $\frac{20}{255}$ ($L_\infty$) and the sum of the absolute value of perturbations over all subpixels is at most $\frac{150}{255}$" ($L_1$ perturbation). While $L_1$ perturbations are not linear (because of the absolute values), reason for which it is seldom used, Prop \ref{prop3} shows how to use it as a perturbation in the MILP model without incurring any expansive binary variables (only cheap linear variables are necessary).

	\item Results obtained on the full input space have several shortcomings: while the bounds are fully correct, they are also particularly pessimistic, as all inputs, including absolutely improbable ones very far away from anything the network has been trained on, need to be accounted for. Also, the runtime to obtain the bound is particularly long. 
	%Finally, when computing worst-case 
	%pairs (image, perturbation), improbable images are generated, hence these  are not meaningful. 
	To solve this problem, we restrict the input space to ignore improbable inputs, using model order reduction techniques from engineering science \cite{aiware}. Specifically, we use Principal Component Analysis (PCA) to reduce the space to a linear input space. We choose the number of dimensions of the space to equal the accuracy of the DNN on the reduced space. 
	%(using a projection to the reduced space then the inverse projection to obtain a very similar image understandable by the DNN). 
	On the MNIST benchmark, this means 20 linear dimensions to match the $97\%$ accuracy of the DNN we considered, instead of the 784 dimensions of the full image space. Using PCA, which is linear, makes it easy to specify perturbations on the actual image (where it is meaningful) rather than on the reduced space.

\item Experimentally, this allows to obtain bounds for a considered pre-trained DNN for MNIST (with accuracy $97\%$) that can certify in real time that $86\%$ of the images are robust for an $L_1$ perturbation of $.5$, resp. $53\%$ for an $L_1$ perturbation of $1$. We finally experimented our MILP models for a {\em regression} task as well, considering a DNN surrogate built from a slow physical finite element model, providing the deformation and the plastic strain of a pipe \cite{aiware}. The DNN has been learnt on a reduced order basis already of dimension 10, from thousands of input dimensions. For regression, the aim is to understand if the DNN surrogate that has been learnt is brittle to small perturbations, which could make it unfaithful to the physical model. Here we want to obtain bounds (which is what our models provide), as well as generating worst-case (image,perturbation) pair, where the strain is particularly different, close to the bound, and that can be run in the physical model to understand whether the DNN surrogate is faithful. Our tool was helpful to provide limited bounds, as well as an almost matching worst-case (image,perturbation) pair that was confirmed in the physical model.
Overall, the "2v" model generates accurate bounds. The decoupling of variables generating the "3v" model offers good trade-offs, reaching slightly lower bounds in the same runtime. 
The "1v" model is faster to converge, but also generates larger bounds, unless runtime is very restricted (Table \ref{table.pipe}).
\end{enumerate}

### 2025-07-31 AAAI2026/notation.tex
global pass

    In this paper, we consider the global verification problem, where we optimize over all image $I$ and all perturbation $I'$ of $I$ with $|I-I'| \leq \varepsilon$. 	We will consider three kinds of variables: 
    \begin{itemize}
    \item $x_j,\hat{x}_j$, for nodes $j$ with input image $I$, 
    \item $x'_j,\hat{x}'_j$, for nodes $j$ with input the perturbed $I'$, 
    \item  $y_j = x_j - x'_j$ the {\em diff variable}, with 
    $\hat{y}_j = \hat{x}_j - \hat{x}'_j$ (and {\em not} 
    $\hat{y}_i = \ReLU(x_j-x'_j)$). 
    \end{itemize}

@@

	\section{(partial) MILP}

@@

	Mixed Integer Linear Programming (MILP) value abstraction is an accurate encoding of ReLU DNNs. For an unstable neuron $n$, that is with values 
    $x \in [\LB(n),\UB(n)]$ with $\LB(n)<0<\UB(n)$, 
    the value $\hat{x}$ of $\ReLU(x)$ can be encoded exactly in an MILP formula with one binary / integer variable $a$ valued in $\{0,1\}$, using constants $\UB(n),\LB(n)$ with 4 constraints \cite{MILP}:

@@

    We use $\mathcal{M}$ to denote the whole MILP model, where each unstable ReLU are encoded in the above way with one integer variable. 
	The encoding from $(\hat{x}_j)_{j \text{ in layer } i}$ to 
	$(x_{j'})_{j' \text{ in layer } i+1}$ variables is simply the linear combination 
	$\boldsymbol{z}^{i} = \boldsymbol{W}^i\cdot \hat{\boldsymbol{z}}^{i-1}+ \vb^i$.
	This exact MILP encoding is often too computationally intensive, as the worst-case complexity of MILP is exponential in the number of integer variables \cite{DivideAndSlide}.

>>>>

    In this paper, we consider the global verification problem, where we optimize over all image $I$ and all perturbation $I'$ of $I$ with $|I-I'| \leq \varepsilon$. 	We will consider three kinds of variables: 
    \begin{itemize}
    \item $x_j,\hat{x}_j$, for nodes $j$ with input image $I$, 
    \item $x'_j,\hat{x}'_j$, for nodes $j$ with input the perturbed $I'$, 
    \item  $y_j = x_j - x'_j$ the {\em diff variable}, with 
    $\hat{y}_j = \hat{x}_j - \hat{x}'_j$ (and {\em not} 
    $\hat{y}_i = \ReLU(x_j-x'_j)$), similarly than in the ITNE model \cite{lipshitz}. 
    \end{itemize}

@@

	%\section{MILP for local robustness}

@@

	Mixed Integer Linear Programming (MILP) can encode faithfully ReLU DNNs:
	For an unstable neuron $n$, that is with values 
    $x \in [\LB(n),\UB(n)]$ with $\LB(n)<0<\UB(n)$, 
    the value $\hat{x}$ of $\ReLU(x)$ can be encoded exactly in an MILP formula with one binary / integer variable $a$ valued in $\{0,1\}$, using constants $\UB(n),\LB(n)$ with 4 constraints \cite{MILP}:

@@

    Let $\mathcal{M}$ denote the whole MILP model, where each unstable ReLU are encoded in the above way with one integer variable. 
	The encoding from $(\hat{x}_j)_{j \text{ in layer } i}$ to 
	$(x_{j'})_{j' \text{ in layer } i+1}$ variables is simply the linear combination 
	$\boldsymbol{z}^{i} = \boldsymbol{W}^i\cdot \hat{\boldsymbol{z}}^{i-1}+ \vb^i$.
	This exact MILP encoding is often too computationally intensive, as the worst-case complexity of MILP is exponential in the number of integer variables \cite{DivideAndSlide}.
